---
title: "AMP-AD1.0 MAYO QC"
author: "Jessica Lundin and ADTR Science Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    embed-resources: true
    self_contained: true
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
#install.packages("pacman")
#library(sageRNAUtils) #in Jaclyn's GitHub repo
#install.packages("knitr")
#install.packages("kableExtra")
pacman::p_load(synapser, dplyr, ggplot2,viridis,patchwork,matrixStats,stringr,forcats,plotly,knitr,kableExtra)
FastQC <- FALSE
MultiQC <- FALSE
```

```{r functions2, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
#setwd("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/functions/functions_demo_plots_from_Jaclyn.R")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/functions/functions_sex_prediction.R")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/functions/functions_pca_outliers.R")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/functions/functions_JB_fastqc_multiqc.R")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/functions/functions_RNA_metric_PCA.R")
```


```{r data-load, echo=FALSE, warning=FALSE, message=FALSE}
# Load metadata and filtered count data
#source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/RNASeq_DE/AMP_AD1/DE_STEP03_QC_preprocessing_prep_sep_by_tissue.R")

invisible(synLogin(silent=TRUE))
md_all_temp <- synapser::synGet('syn76141636') 
  md_all <- read.csv(md_all_temp$path)
md_all <- md_all[md_all$specimenID != "742_CER",]

counts_tcx_temp <- synapser::synGet('syn76141793') 
  counts_tcx <- read.csv(counts_tcx_temp$path, sep="\t",check.names = FALSE)
  counts_tcx <- counts_tcx[,-2]
counts_cer_temp <- synapser::synGet('syn76141809') 
  counts_cer <- read.csv(counts_cer_temp$path, sep="\t",check.names = FALSE)
  counts_cer <- counts_cer[,-2]

counts <- merge(counts_cer, counts_tcx, by="gene_id")  
  counts_rn <- merge(counts_cer, counts_tcx, by="gene_id")  
meta_age <- md_all |>
  mutate(ageDeath = case_when(
    is.na(ageDeath) | ageDeath == "missing or unknown" ~ NA,
    ageDeath == "90+" ~ 90,
    .default = suppressWarnings(as.numeric(ageDeath))
  )
)

  # Align samples - TO DO 
  samples  <- intersect(colnames(counts), md_all$specimenID)
  if (length(samples) == 0) stop("No overlapping samples between counts and metadata.")
  counts  <- counts[, samples, drop = FALSE]
  md_all <- md_all[match(samples, md_all$specimenID), , drop = FALSE]
  rownames(md_all) <- md_all$specimenID
  rownames(counts) <- counts_rn$gene_id
  
#metadata <- md_all
```

# Definitions
**Abbreviations:** CEB = cerebellum, TCX = temporal cortex

**AD diagnosis:** Based on Mayo neurologist diagnosis. 

**Control diagnosis:** Based on Mayo neurologist diagnosis. 

Any donors who did not fall under diagnosis or control criteria were assigned as “other”.

n=54 samples excluded based on 'EXCLUDE = TRUE' from Mayo. Reasons include GeneBodyCoverage Plot (RSeQC) (n=19), Heterozygosity Outlier (n=9), Principal Components Analysis (n=5), Sex mis-match (n=9), 3rd degree relative with another sample (n=4), Eigenstrat Outlier (n=3), Does not meet control criteria (Braak > 3.0) (n=4), Low Mapped Reads (n=1)

After the exclusion above and a restriction to those with available transcriptomic data, there were n=160 with a diagnosis of Alzheimer's Disease, n=53 with a diagnosis of "pathological aging", n=161 with a diagnosis of "progressive supranuclear palsy", and n=134 "control" participants.

The final dataset included n=160 AD and n=134 controls

**Starting dataset: n=`r nrow(md_all)` samples**

# Descriptive: Demographics
```{r functions, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/RNASeq_DE/AMP_AD1/functions/functions_demo_plots_from_Jaclyn.R")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/RNASeq_DE/AMP_AD1/functions/functions_sex_prediction.R")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/RNASeq_DE/AMP_AD1/functions/functions_pca_outliers.R")
source("C:/Users/jlundin/OneDrive/git_code/RNASeq_DE/AMP_AD1/RNASeq_DE/AMP_AD1/functions/functions_JB_fastqc_multiqc.R")
```


```{r generate-demographics-plots, warning=FALSE, message=FALSE, fig.width=12, fig.height=17}
## Generate QC plots

md_all$ADdiagnosis <- md_all$diagnosis
p1<-make_bar_plot(md_all, "tissue2", facet_var = NULL)
p2<-make_bar_plot(md_all, "ADdiagnosis")
p3<-make_bar_plot(md_all, "sex")
p4<-make_bar_plot(md_all, "race")
p5<-make_bar_plot(md_all, "isHispanic")
p6<-make_bar_plot(md_all, "cohort")
p7<-(ggplot(meta_age, aes(x = ageDeath)) + labs(title = "ageDeath")) |>
  format_histogram_plot()
p8<-make_bar_plot(md_all, "age_cat", facet_var = "diagnosis")


library(patchwork)
plot_list <- list(p1, p2, p3, p4, p5, p6, p7, p8)
wrap_plots(plot_list, nrow = 4, ncol = 2)

```

# Descriptive: Technical covariates

number of missing for PMI: `r length(md_all[is.na(md_all$PMI),])` 

number of missing values for RIN: `r length(md_all[is.na(md_all$RIN),])` 

```{r generate-technical-covariate-plots, warning=FALSE, message=FALSE, fig.width=12, fig.height=5}
#make_bar_plot_batch(md_all, "procbatch")
#make_bar_plot_batch(md_all, "libraryPrep")
#make_bar_plot_batch(md_all, "final_batch")
#make_bar_plot_batch_fillvar(md_all, "final_batch", facet_var = "tissue", fill_var = "procbatch")
#make_bar_plot_batch_fillvar(md_all, "final_batch", facet_var = "tissue", fill_var = "libraryPrep")

p9<-(ggplot(md_all, aes(x = PMI)) + labs(title = "PMI")) |>
  format_histogram_plot()

p10<-(ggplot(md_all, aes(x = RIN)) + labs(title = "RIN")) |>
  format_histogram_plot()

library(patchwork)
plot_list <- list(p9, p10)
wrap_plots(plot_list, nrow = 1, ncol = 2)
```

```{r generate-RIN-plots1, warning=FALSE, message=FALSE}
md_all_RIN_na <- md_all[is.na(md_all$RIN),] 
md_all_RIN_WARN <- md_all[(is.na(md_all$RIN) | md_all$RIN < 2.5),] 
```

# RIN
**Samples were marked as QC failures if RIN was NA (n=`r nrow(md_all_RIN_na)`)**

**Samples were flagged as QC warning if RIN < 2.5 (n=`r nrow(md_all_RIN_WARN)`)**

RIN is a numerical value that indicates RNA integrity. 

```{r generate-RIN-plots2, warning=FALSE, message=FALSE}
md_all_RIN_na <- md_all[is.na(md_all$RIN),] 
md_all_RIN_WARN <- md_all[(is.na(md_all$RIN) | md_all$RIN < 2.5),] 

plt1 <- ggplot(md_all[!is.na(md_all$RIN),] , aes(x = tissue2, y = RIN, fill = tissue2)) +
  geom_boxplot(outliers = FALSE, na.rm = TRUE, width = 0.5) +
  geom_jitter(aes(color=RIN<2.5), size = 0.5, na.rm = TRUE, width = 0.3) +
  scale_color_manual(values=c("TRUE" = "red", "FALSE" = "black")) +
  xlab(NULL) +
  theme_bw() +
  scale_fill_viridis(discrete = TRUE, begin = 0.1, end = 0.9, alpha = 0.7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        title = element_text(face = "bold")) +
  labs(color = "RIN < 2.5", fill = "Tissue")

print(plt1)

## table
#tab_data <- as.data.frame.matrix(table(md_all_RIN_WARN$tissue2))

#tab_data %>% kable(format = "html", caption = "RIN < 2.5 or missing RIN data: Sample count by tissue and final_batch") %>%
#  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
#                full_width = F, position = "left") %>%
#  row_spec(0, bold = TRUE, color = "white", background = "#444444")

```


# Inferred sex

Examine average expression of key Y-chromosome genes to infer the sex of each sample. Genes used are RPS4Y1, EIF1AY, DDX3Y, and KDM5D, which were determined to be robust markers for sex in Staedtler, et al (2013). 

Samples whose inferred sex did not match their reported sex were marked as QC failures (for now).

```{r generate-sex prediction-plots1, warning=FALSE, message=FALSE}

## Generate sex prediction plots
# inputs 
#sc_data <- rosmap_reads_DLPFC
#md_brain_tissue <- md_DLPFC
#tissue <- "DLPFC"


#tissue2_var <- "DLPFC"
#tissue2_var <- "CN"
#for (j in 1:length(tissue2_var)){

sc_data <- counts
#md_brain_tissue <- md_all[md_all$tissue2 == tissue2_var,]
md_brain_tissue <- md_all
tissue <- "all"

# gene expression reads file
sc_data$gene_id <- rownames(sc_data)
sc_data$gene_id <- sub("\\..*$", "", sc_data$gene_id)
sc_data2 <- sc_data[!duplicated(sc_data$gene_id),]
sc_data2 <- as.data.frame(sc_data2)
rownames(sc_data2) <- sc_data2$gene_id
sc_data2 <- sc_data2 %>% dplyr::select(-gene_id)

#specID_tissue <- md_all$specimenID[md_all$tissue2 == tissue2_var]
#sc_data2 <- sc_data2[,colnames(sc_data2) %in% specID_tissue]


sex_pred_strat = FALSE
if(sex_pred_strat){
#delete_var <- read.csv(synapser::synGet('syn75279942')$path, sep="\t", stringsAsFactors = F, check.names = FALSE)
#delete_var2 <- delete_var[delete_var$tissue2 == tissue2_var,]
sc_data2_filtered <- sc_data2
sex_var <- c("female","male")
for (i in 1:length(sex_var)){
# 1. Identify the specific genes and specimens to target
#target_genes <- delete_var2$gene_id_clean[delete_var2$sex == sex_var[i]]
target_genes <- delete_var2$gene_id_clean
target_specs <- md_brain_tissue$specimenID[md_brain_tissue$sex == sex_var[i]]

# 2. Filter down to only those that actually exist in your matrix
valid_genes <- intersect(target_genes, rownames(sc_data2))
valid_specs <- intersect(target_specs, colnames(sc_data2))

# 3. Create a coordinate grid matrix of row/column combinations
coord_matrix <- expand.grid(row = valid_genes, col = valid_specs, stringsAsFactors = FALSE)
coord_matrix <- as.matrix(coord_matrix)

# 4. Turn those specific cells into NA (representing NULL)
#sc_data2[coord_matrix] <- NA
# 1. Convert your data frame to a matrix
sc_matrix <- as.matrix(sc_data2_filtered)

# 2. Apply your coordinates (this will work now!)
sc_matrix[coord_matrix] <- NA

# 3. Convert it back to a data frame
sc_data2_filtered <- as.data.frame(sc_matrix)
}

#sc_data2[1:5,1:5]
expr_mat <- sc_data2_filtered
sc_data2 <- sc_data2_filtered
}

# md file with self reported sex
md_brain_tissue_sorted <- md_brain_tissue [match(colnames(sc_data2), md_brain_tissue$specimenID), ]
reported_sex_vector <- md_brain_tissue_sorted$sex

# Vector of names
names <- colnames(sc_data2)

# Assign names
names(reported_sex_vector) <- names

expr_mat <- sc_data2 

# run code 
result <- predict_sex(expr_mat, reported = reported_sex_vector, threshold = 0)
result_md <- merge(result, md_brain_tissue, by.x="sample", by.y="specimenID", all=T)
result_md_filtered <- result_md  %>% group_by(individualID) %>%
  filter(any(concordant == FALSE)) %>% ungroup()

result2 <- result %>%
  left_join(md_brain_tissue_sorted %>% dplyr::select(specimenID, tissue2, individualID), by = c("sample" = "specimenID"))

#attr(sc_data2, "threshold") <- list(xist = 2^10, y = 2^4)
#attr(sc_data2, "threshold") <- list(xist = 2^10)
#thr <- attr(sc_data2, "threshold")
  
result2 <- result2 |> mutate(concordant2 = case_when(
     !concordant                                  ~ "Discordant",
      TRUE                                         ~ "Concordant" ))

QC_failures_sex <- result2 %>% filter(concordant2 == "Discordant"  )
QC_nonfailures_sex <- result2 %>% filter(concordant2 == "Concordant" )
    
meta_sub_sex_unique <- result2 %>% filter(concordant2 == "Discordant" ) %>% distinct(individualID, .keep_all = TRUE)

```
**Reported sex vs inferred sex mismatches: `r nrow(QC_failures_sex)` sample(s) from `r nrow(meta_sub_sex_unique)` individuals across all brain tissue regions (n=`r ncol(sc_data2)` total samples).** 

```{r generate-sex prediction-plots2, warning=FALSE, message=FALSE}
#plot_sex_scores(result_md)           # XIST vs Y-score scatter
plot_sex_scores(result_md)  


## table
meta_sub_sex <- result2 %>% filter(concordant2 == "Discordant") |>
  dplyr::select(sample, tissue2, concordant2) |>
  dplyr::rename(`Specimen ID` = sample,
                Tissue = tissue2,
                QC = concordant2) |>
  arrange(QC, Tissue, `Specimen ID`)

# Create the table
kable(meta_sub_sex, format = "html", caption = "Reported sex vs inferred sex mismatches") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444")

```


```{r generate-PCA-plots, fig.width=10, fig.height=3, warning=FALSE, message=FALSE, eval=FALSE}
# PCA outliers

Samples were split by tissue and batch. PCA was run on each group, and outliers were defined as samples falling outside 4 standard deviations of the data along PC1, PC2, and/or PC3. 

**NOTE:** circles are samples than pass QC, triangles are samples that fail QC. "final_batch" in the plot header is: processing batch and sequencing batch separated by an underscore. Dashed red rectangle shows outlier thresholds. Only batches with >10 samples were analyzed for outliers using this method.



#delete_var <- read.csv(synapser::synGet('syn75279942')$path, sep="\t", stringsAsFactors = F, check.names = FALSE)
#combined_full_all4 <- as.data.frame(combined_full_all4)
outlier_list <- list()
tissue2 <- unique(md_all$tissue2)
for (i in 1:length(tissue2)){    
  if (tissue2[i] == "TCX"){
    md_brain_tissue <- md_all %>% dplyr::filter(tissue2 == "TCX")
    tissue <- "TCX"
    tissue2_var <- "TCX"
      }
    if (tissue2[i] == "CER"){
    md_brain_tissue <- md_all %>% dplyr::filter(tissue2 == "CER")
    tissue <- "CER"
    tissue2_var <- "CER"
      }

   # rownames(counts) <- counts$gene_id
    sc_data <- counts[-1] %>% dplyr::select(any_of(md_brain_tissue$specimenID))
   # batch_loop <- md_brain_tissue %>% filter(n() > 10) %>% ungroup()      
    batch_loop <- 1
    
for (j in 1:length(batch_loop)){    
  #  md_brain_tissue_loop <- md_brain_tissue[md_brain_tissue$final_batch == batch_loop [j],]
  #  spec_var <- md_brain_tissue_loop$specimenID
 #   sc_data_loop <- sc_data %>% dplyr::select(all_of(spec_var))

    
  #  pca_result <- run_pca(sc_data_loop, md_brain_tissue_loop, tissue)
  pca_result <- run_pca(sc_data, md_brain_tissue, tissue_var = tissue)
    plot_pca_panels(pca_result, colour_var = NULL )
    #plot_pca(pca_result, "PC1", "PC2", "final_batch" )
    QC_failures_pca_x <- pca_result$scores %>% dplyr::select(specimenID, is_outlier) 

    # This filters the data and pulls just the specimenIDs into a vector/list
    new_outlier_list <- QC_failures_pca_x %>%
      filter(is_outlier == TRUE) %>%
      pull(specimenID)
    outlier_list <- append(outlier_list, new_outlier_list)
  }
}

## table
pca_final <- md_all %>% filter(specimenID %in% outlier_list) %>% dplyr::select(specimenID, tissue, individualID)
            
pca_final2 <- pca_final %>%
  dplyr::select(specimenID, tissue) |>
  arrange(tissue, specimenID)

**Final PCA outliers, n=`r length(outlier_list)`**
```


```{r PCA table, warning=FALSE, message=FALSE, eval=F}
# Create the table
kable(pca_final2, format = "html", caption = "PCA outliers by tissue") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444")

```


# Validation for FastQC

Examine Phred score and average base content at each position of a read.

Samples fail QC if the Phred score is < 20.

Samples get a QC warning if the Phred score is below Q1 - 1.5 * IQR.

Samples get a QC warning if the base content is above Q3 + 1.5 * IQR for any base. 

Samples with 2 or more warnings across all QC metrics fail QC.


```{r fast-qc validation, warning=FALSE, message=FALSE}
# This code will ONLY run if FastQC is TRUE.
# If FastQC is FALSE, the code is skipped entirely during knitting.

# Load fastqc stats (Mayo_fq_stats.rds)
invisible(synLogin(silent=TRUE))
file_entity3 <- synGet("syn76818030") #.rds "basic_stats", "phred_per_base", "base_content"
load(file_entity3$path)
basic_stats <- Mayo_bundle$basic_stats
phred_per_base <- Mayo_bundle$phred_per_base
base_content <- Mayo_bundle$base_content

# Load gene metadata
syn_gene_file <- synapser::synGet('syn74907403')
gene_info <- read.csv(syn_gene_file$path)

base_content2 <- merge(base_content, md_all, by="specimenID")
basic_stats2 <- merge(basic_stats, md_all, by="specimenID")
phred_per_base2 <- merge(phred_per_base, md_all, by="specimenID")


# validate-fastqc 
thresholds <- NULL
thresholds$phred <- 20

base_content2 <- base_content2 |>
  group_by(read, tissue2, base) |>
  mutate(outlier_color = ifelse(is_outlier_IQR(mean_base_deviation),
                                "black", NA)) |>
  ungroup()

plt1b <- ggplot(base_content2,
               aes(x = base, y = mean_base_deviation, fill = base)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(aes(color = outlier_color), # <-- Map inside aes() without using $
              size = 0.5, width = 0.05, na.rm = TRUE) +
  scale_color_identity() +                # <-- Tells ggplot to use the actual color strings
  theme_bw() +
  facet_grid(rows = vars(read), cols = vars(tissue2)) +
  scale_fill_viridis(discrete = TRUE, begin = 0.1, alpha = 0.8) +
  theme(strip.background = element_rect(fill = "#EEEEEE", color = "black"),
        strip.text = element_text(face = "bold"),
        title = element_text(face = "bold")) +
  ggtitle("Mean deviation from expected base proportions")

print(plt1b)

# Using the mean content deviation from 25% of each base at each position.
# Mayo uses sum of deviation instead of mean, but they're equivalent for the
# purposes of outlier finding. Unlike for the graph, we only want outliers that
# are on the upper tail (over-representation of a base).
base_content_outliers <- base_content2 |>
  dplyr::group_by(tissue, read, base) |>
  mutate(is_outlier = is_outlier_IQR(mean_base_deviation, tail = "upper")) |>
  subset(is_outlier == TRUE) |>
  pull(specimenID)

# Any base falling below the Phred threshold marks the sample as failing QC
phred_fail <- phred_per_base |>
  subset(Median < thresholds$phred) |>
  pull(specimenID)

phred_outliers <- phred_per_base |>
  group_by(position) |>
  mutate(is_outlier = is_outlier_IQR(Median, tail = "lower")) |>
  subset(is_outlier == TRUE) |>
  pull(specimenID)

metadata <- md_all
metadata$base_content_warn <- metadata$specimenID %in% base_content_outliers
metadata$phred_score_valid <- !(metadata$specimenID %in% phred_fail)
metadata$phred_score_warn <- metadata$specimenID %in% phred_outliers

n_fqc_warn_fail <- sum(metadata$base_content_warn | metadata$phred_score_warn |
                         !metadata$phred_score_valid)


# ---- print_fastqc-results
meta_sub_fastqc <- subset(metadata, base_content_warn | phred_score_warn | !phred_score_valid) |>
  dplyr::select(specimenID, tissue, base_content_warn, phred_score_warn, phred_score_valid) |>
  dplyr:: mutate(
    `Phred Score` = case_when(
      !phred_score_valid ~ "FAIL",
      phred_score_warn ~ "WARN",
      .default = "."
    ),
    `Base Content` = ifelse(base_content_warn == TRUE, "WARN", ".")
  ) |>
  dplyr::select(specimenID, tissue, `Phred Score`, `Base Content`) |>
  dplyr::rename(Tissue = tissue, `Specimen ID` = specimenID) |>
  arrange(Tissue, `Specimen ID`)

#cat("This table shows Phred QC warnings vs Base Content QC warnings -- FAIL if QC warning for both?")
#print(table(meta_sub_fastqc$`Phred Score`, meta_sub_fastqc$`Base Content`))


# Move this back out when pastqc is run
#**FastQC Phred failure(s):** `r sum(!metadata$phred_score_valid)` sample(s)

#**FastQC Phred warning(s):** `r sum(metadata$phred_score_warn)` sample(s)

#**FastQC base content warning(s):** `r sum(metadata$base_content_warn)` sample(s)
```



# Validation for MultiQC

Examined percentage of reads mapped and percentage of reads duplicated.

Samples were marked as QC failure if < 70% reads mapped or > 80% duplicated reads.

Samples were flagged as QC warning if reads mapped reported less than Q1 - 1.5 * IQR or duplicate reads reported more than Q3 + 3 * IQR. QC warnings were evaluated independently for each brain region (tissue type).

```{r multi-qc validation, echo=FALSE, warning=FALSE, message=FALSE}
reads_mapped <- 70
reads_duplicated <- 80

reads_mapped_fail <- md_all |>
  subset(samtools_reads_mapped_percent < reads_mapped) |>
  pull(specimenID)


tissue_var <- unique(md_all$tissue)
reads_mapped_outliers <- purrr::map(tissue_var, \(t) {
 md_all |>
    filter(tissue == t) |>
    mutate(is_outlier = is_outlier_IQR(samtools_reads_mapped_percent, tail = "lower")) |>
    filter(is_outlier == TRUE) |>
    pull(specimenID)
}) |>
  unlist()

md_all$picard_PERCENT_DUPLICATION <- md_all$picard_PERCENT_DUPLICATION * 100
reads_dupe_fail <- md_all  |>
  subset(picard_PERCENT_DUPLICATION > reads_duplicated) |>
  pull(specimenID)

# Use Q3 + 3*IQR for outliers here
#reads_dupe_outliers <- multiqc_stats |>
#  mutate(is_outlier = is_outlier_IQR(picard_PERCENT_DUPLICATION,
#                                     tail = "upper", IQR_mult = 3)) |>
#  subset(is_outlier == TRUE) |>
#  pull(specimenID)

reads_dupe_outliers <- purrr::map(tissue_var, \(t) {
  md_all  |>
    filter(tissue == t) |>
    mutate(is_outlier = is_outlier_IQR(picard_PERCENT_DUPLICATION,
                                       tail = "upper", IQR_mult = 3)) |>
    filter(is_outlier == TRUE) |>
    pull(specimenID)
}) |>
  unlist()


md_all$reads_mapped_fail <- md_all$specimenID %in% reads_mapped_fail
md_all$reads_mapped_warn <- md_all$specimenID %in% reads_mapped_outliers
md_all$reads_mapped_valid <- !(md_all$specimenID %in% reads_mapped_fail)

md_all$reads_duplicated_fail <- md_all$specimenID %in% reads_dupe_fail
md_all$reads_duplicated_warn <- md_all$specimenID %in% reads_dupe_outliers
md_all$reads_duplicated_valid <- !(md_all$specimenID %in% reads_dupe_fail)

#mqc_plot <- multiqc_stats |>
#  mutate(
#    mapped_status = case_when(
#      specimenID %in% reads_mapped_fail ~ "Fail",
#      specimenID %in% reads_mapped_outliers ~ "Warn", 
#      .default = "Pass"
#    ),
#    duplicated_status = case_when(
#      specimenID %in% reads_dupe_fail ~ "Fail",
#      specimenID %in% reads_dupe_outliers ~ "Warn", 
#      .default = "Pass"
#    )
#  )

mqc_plot <- md_all  |>
  mutate(
    mapped_status = case_when(
      specimenID %in% reads_mapped_fail ~ "Fail",
      specimenID %in% reads_mapped_outliers & !specimenID %in% reads_mapped_fail ~ "Warn",
      .default = "Pass"
    ),
    duplicated_status = case_when(
      specimenID %in% reads_dupe_fail ~ "Fail",
      specimenID %in% reads_dupe_outliers & !specimenID %in% reads_dupe_fail ~ "Warn",
      .default = "Pass"
    )
  ) |>
  mutate(mapped_status = factor(mapped_status, levels = c("Fail", "Warn", "Pass")))
#levels(mqc_plot$mapped_status)


stat_colors <- c("Pass" = "black", "Warn" = "grey75", "Fail" = "red3")
plt1 <- ggplot(mqc_plot,
               aes(x = tissue2, y = samtools_reads_mapped_percent, fill = tissue2)) +
  geom_boxplot(outliers = FALSE, width = 0.1) +
  geom_jitter(aes(color = mapped_status),
              width = 0.2,
              size = ifelse(mqc_plot$mapped_status == "Pass", 0.5, 1)) +
  theme_bw() +
  theme(legend.position = "none",
        title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab(NULL) +
  scale_color_manual(values = stat_colors, limits = c("Fail", "Warn", "Pass")) +
  scale_fill_viridis(name = "Tissue", discrete = TRUE, begin = 0.2, alpha = 0.7) +
  ggtitle("Percentage of reads mapped")

plt2 <- ggplot(mqc_plot,
               aes(x = tissue2, y = picard_PERCENT_DUPLICATION, fill = tissue2)) +
  geom_boxplot(outliers = FALSE, width = 0.1) +
  geom_jitter(aes(color = duplicated_status),
              width = 0.2,
              size = ifelse(mqc_plot$duplicated_status == "Pass", 0.5, 1)) +
  theme_bw() +
  theme(title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab(NULL) +
  labs(color = "Status") +
  scale_color_manual(values = stat_colors, limits = c("Fail", "Warn", "Pass")) +
  scale_fill_viridis(name = "Tissue", discrete = TRUE, begin = 0.2, alpha = 0.7) +
  ggtitle("Percentage of reads duplicated") +
  theme(plot.margin = unit(c(0, 0, 0, 50), units = "pt"))

print(plt1 + plt2)

n_mqc_fail <- sum(md_all$reads_mapped_fail |md_all$reads_duplicated_fail |
                         !md_all$reads_mapped_valid | !md_all$reads_duplicated_valid)
n_mqc_warn <- sum(md_all$reads_mapped_warn |md_all$reads_duplicated_warn |
                         !md_all$reads_mapped_valid | !md_all$reads_duplicated_valid)


# add back in later
#**MutliQC failures:** `r n_mqc_fail` sample(s)

#**MutliQC warnings:** `r n_mqc_warn` sample(s)

```



```{r multiqc-table-failures, warning=FALSE, message=FALSE}
meta_sub_warn <- subset(md_all, reads_mapped_warn | reads_duplicated_warn |
                     !reads_mapped_valid | !reads_duplicated_valid) |>
  dplyr::select(specimenID, tissue, tissue2, reads_mapped_warn, reads_duplicated_warn,
         reads_mapped_valid, reads_duplicated_valid) |>
  dplyr::mutate(
    `Reads Mapped` = case_when(
      !reads_mapped_valid ~ "FAIL",
      reads_mapped_warn ~ "WARN",
      .default = "."
    ),
    `Reads Duplicated` = case_when(
      !reads_duplicated_valid ~ "FAIL",
      reads_duplicated_warn ~ "WARN",
      .default = "."
    )
  ) |>
  dplyr::select(specimenID, tissue2, `Reads Mapped`, `Reads Duplicated`) |>
  dplyr::rename(Tissue = tissue2) |>
  dplyr::arrange(Tissue, specimenID)

meta_sub_fail <- subset(md_all, reads_mapped_fail | reads_duplicated_fail |
                          !reads_duplicated_valid |!reads_duplicated_warn) |>
  dplyr::select(specimenID, tissue, tissue2, reads_mapped_fail, reads_duplicated_fail,
        reads_mapped_valid, reads_duplicated_valid) |>
  dplyr::mutate(
    `Reads Mapped` = case_when(
      reads_mapped_fail ~ "FAIL",
    # !reads_mapped_valid ~ "WARN",
      .default = "."
    ),
    `Reads Duplicated` = case_when(
      reads_duplicated_fail ~ "FAIL",
     # !reads_duplicated_valid ~ "WARN",
      .default = "."
    )
  ) |>
  dplyr::select(specimenID, tissue2, `Reads Mapped`, `Reads Duplicated`) |>
  dplyr::filter(`Reads Mapped` != "." | `Reads Duplicated` != ".") |>
  dplyr::rename(Tissue = tissue2, `Specimen ID` = specimenID) |>
  dplyr::arrange(Tissue, `Specimen ID`)

## Create the table for failures by sample
kable(meta_sub_fail, format = "html", caption = "Samples not meeting MultiQC thresholds (failed QC)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444")
## Create a table for warnings grouped by tissue type
# 1. Group by tissue and "collapse" the IDs into a single string
tab_summary2 <- meta_sub_warn %>%
  group_by(Tissue) %>%
  summarize(
    Count = n(),
    # 'collapse' is the magic argument that turns a list into a comma-separated string
    SpecimenIDs = paste(specimenID, collapse = ", ")
  )

# 2. Format as a clean table
tab_summary2 %>%
  kable(format = "html", caption = "Samples flagged as MultiQC warning by Specimen IDs and Tissue Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444") %>%
  # Crucial: Set a width for the long ID column so it wraps text
  column_spec(3, width = "15cm")

```



# Final samples that passed QC

```{r saving-rdata-final-QC-tables-part1-tablesetup, warning=FALSE, message=FALSE}

sex_fail <- meta_sub_sex$`Specimen ID`[meta_sub_sex$QC == "Discordant"]
sex_warn <- meta_sub_sex$`Specimen ID`[meta_sub_sex$QC == "Flagged"]
#pca_fail <- pca_final2$specimenID
RIN_fail <- md_all_RIN_na$specimenID
RIN_warn <- md_all_RIN_WARN$specimenID
multiqc_fail <- meta_sub_fail$`Specimen ID`
multiqc_warn <- meta_sub_warn$specimenID
fastqc_phred_fail <- meta_sub_fastqc$`Specimen ID`[meta_sub_fastqc$`Phred Score` == "FAIL"]
fastqc_phred_warn <- meta_sub_fastqc$`Specimen ID`[meta_sub_fastqc$`Phred Score` == "WARN"]
fastqc_base_warn <- meta_sub_fastqc$`Specimen ID`[meta_sub_fastqc$`Base Content` == "WARN"]

exclude_ids <- unique(c(RIN_fail, multiqc_fail, sex_fail, sex_warn, fastqc_phred_fail  ))
exclude_ids_warn <- unique(c(multiqc_warn, RIN_warn, fastqc_phred_warn, fastqc_base_warn))

#exclude_ids <- unique(c(pca_fail, RIN_fail, sex_fail, sex_warn ))
#exclude_ids_warn <- unique(c(RIN_warn))


md_valid <- md_all %>%
  mutate(
    valid = case_when(
      specimenID %in% exclude_ids      ~ "FAIL",
      specimenID %in% exclude_ids_warn ~ "WARN",
      TRUE                             ~ "PASS"  # "TRUE" acts as the 'else' catch-all
    )
  )
#table(md_valid$valid, useNA="always")

n_passes <- table(md_valid$tissue2, md_valid$valid)
colnames(n_passes) <- c("Pass", "Warn")

failures <- subset(md_valid, valid == "FAIL")
warn <- subset(md_valid, valid == "WARN")
fail_warn <- subset(md_valid, valid == "FAIL" | valid == "WARN")
md_valid_qc <- subset(md_valid, valid == "PASS" | valid == "WARN")

## saving .rds with QC'd data
counts <- counts[, md_valid_qc$specimenID]

#multiqc_stats <- subset(multiqc_stats, specimenID %in% md_valid$specimenID)
#fastqc_basic_stats <- subset(fastqc_data$basic_statistics, specimenID %in% md_valid$specimenID)

# Remove genes that are all 0's
zeros <- rowSums(counts) == 0
#table(zeros)

```
**`r nrow(md_valid_qc)` of `r nrow(md_all)` samples passed QC.** 

```{r final-QC-tables-part2-crosstabs-and-summary, warning=FALSE, message=FALSE}

tab_data2 <- as.data.frame.matrix(table(md_valid$tissue2, md_valid$valid))
tab_data2 %>% kable(format = "html", caption = "Samples by QC status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, position = "left") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444")

# 1. Group by tissue and "collapse" the IDs into a single string
tab_summary <- failures %>%
  group_by(tissue2) %>%
  summarize(
    Count = n(),
    # 'collapse' is the magic argument that turns a list into a comma-separated string
    SpecimenIDs = paste(specimenID, collapse = ", ")
  )

# 2. Format as a clean table
tab_summary %>%
  kable(format = "html", caption = "Samples that failed QC by Tissue Type and Specimen ID",
                col.names = c("Tissue", "Count", "Specimen ID")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444") 

```

``` {r final-QC-tables-part3-details-failed-QC, warning=FALSE, message=FALSE}

#failures_detail <- fail_warn %>%
#  dplyr::select(specimenID, tissue2, valid) %>%
#  dplyr::mutate(
#    `Sex check` = ifelse(fail_warn$specimenID %in% sex_fail, "FAIL", "."),
#    `PCA check` = ifelse(fail_warn$specimenID %in% pca_fail, "FAIL", "."),
#    `RIN check` = ifelse(fail_warn$specimenID %in% RIN_fail, "FAIL", "."),
#    `RIN < 2.5` = ifelse(fail_warn$specimenID %in% RIN_warn, "WARN", "."),
#    `MultiQC fail` = ifelse(fail_warn$specimenID %in% multiqc_fail, "FAIL", "."),
#    `MultiQC warn` = ifelse(fail_warn$specimenID %in% multiqc_warn, "WARN", ".")) %>%
#  dplyr::select(specimenID, tissue2, `Sex check`, `PCA check`, `RIN check`, `RIN < 2.5`, `MultiQC fail`, `MultiQC warn) %>%
#  dplyr::rename(Tissue = tissue2) %>%
#  dplyr::arrange(Tissue, specimenID)

failures_detail2 <- fail_warn |>
  dplyr::select(specimenID, tissue2, valid) |>
  dplyr::mutate(
    `Sexfail` = ifelse(specimenID %in% sex_fail, "FAIL", "."),
    `Sexwarn` = ifelse(specimenID %in% sex_warn, "FAIL", "."),
   # `PCAfail` = ifelse(specimenID %in% pca_fail, "FAIL", "."),
    `RINmissing` = ifelse(specimenID %in% RIN_fail, "FAIL", "."),
    `MultiQCfail` = ifelse(specimenID %in% multiqc_fail, "FAIL", "."),
    `MultiQCwarn` = ifelse(specimenID %in% multiqc_warn, "WARN", "."),
    `FastQCfail` = ifelse(specimenID %in% fastqc_phred_fail, "FAIL", "."),
    `FastQCwarn1` = ifelse(specimenID %in% fastqc_phred_warn, "WARN", "."),
    `FastQCwarn2` = ifelse(specimenID %in% fastqc_base_warn, "WARN", ".")
  ) |>
  dplyr::select(specimenID, tissue2, `Sexfail`, `Sexwarn`,  `RINmissing`, `MultiQCfail`,`MultiQCwarn`,`FastQCfail`,`FastQCwarn1`,`FastQCwarn2`) |>
  dplyr::filter(if_any(c(`Sexfail`, `Sexwarn`, `RINmissing`, `MultiQCfail`,`FastQCfail`), ~ .x == "FAIL") | if_any(c(`Sexwarn`), ~ .x == "FLAG")) |>
  dplyr::rename(Tissue = tissue2) |>
  arrange(Tissue, specimenID)


failures_detail2 %>%
  kable(format = "html", caption = paste("Details of why each of ", nrow(failures_detail2), " sample(s) failed QC. Added MultiQC and FastQC warning flags to existing QC failures, but these samples were not not excluded. Sample flagged as outliers for inferred sex were excluded, but shown separately from discordant failed samples.", sep=""),
        col.names = c("Specimen ID", "Tissue", "Sex fail", "Sex outlier", 
                  "RIN missing", "MultiQC fail", "MultiQC warn", "FastQC Phred fail",  
                  "FastQC phred warn", "FastQC base content warn")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
              fixed_thead = TRUE) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444") %>%
  scroll_box(width = "100%", height = "500px")

```


# Metadata and clean counts

Uploaded to synapse: "MAYO_md_counts.rds", parent = "syn75817798"

*`r nrow(md_valid_qc)` sample(s)*

*`r nrow(counts)` genes(s)*

```{r write synapse rds, eval=FALSE}
data_final <- list("metadata" = md_valid_qc, "counts" = counts[!zeros, ])
                   #"multiqc_stats" = multiqc_stats,
                   #"fastqc_stats" = fastqc_basic_stats)
output_filename <- "MAYO_md_counts.rds"
saveRDS(data_final, file = output_filename)
synapse_file_object <- File(path = "MAYO_md_counts.rds", parent = "syn75817798")
synStore(synapse_file_object)
```


# CQN normalization

Final cqn outputs uploaded to synapse: "MAYO_md_counts_cqn.rds", parent = "syn75817798"


**Library preparation bias:** Library preparation method is a systematic technical covariate that affects GC content and gene length biases because these curves are not constant across library prep methods.

```{r cqn, fig.width=12, fig.height=10, echo=FALSE, warning=FALSE}
invisible(synLogin(silent=TRUE))
file_entity <- synGet("syn76747558") # MAYO_qc.rds "metadata" and "counts" 
invisible(list2env(readRDS(file_entity$path), envir = .GlobalEnv))

#dim(counts)
md <- metadata


library(cqn)
# Align samples 
syn_gene_file <- synapser::synGet('syn74907403')
gene_info_path <- read.csv(syn_gene_file$path)
gene_info_path <- gene_info_path %>% dplyr::rename(gene_id = ensembl_gene_id, gc_content = percent_gc_content, length = gene_length)
rownames(gene_info_path) <- gene_info_path$gene_id

gene_var  <- intersect(rownames(counts), gene_info_path$gene_id)
gene_info_path2 <- gene_info_path[gene_info_path$gene_id %in% gene_var,]
rownames(gene_info_path2) <- gene_info_path2$gene_id

cqn_fit <- cqn(
  counts     = counts,
  x          = gene_info_path2$gc_content,      # per-gene GC fraction
  lengths    = gene_info_path2$length,         # per-gene length
  sizeFactors = colSums(counts))

cqnplot(cqn_fit, n = 1, xlab = "GC content", ylab = "QR fit", main=paste("CQN"))
cqnplot(cqn_fit, n = 2, xlab = "Length (log2)", ylab = "QR fit", main=paste("CQN"))

# CQN offset = glm.offset (combined log-scale offset)
cqn_offsets <- cqn_fit$glm.offset 
cqn_E <- cqn_fit$y + cqn_fit$offset 

```


```{r cqn function with output, eval=TRUE, warning=FALSE, include=FALSE}
# -- 1. Incorporate into limma-voom
# ── 2. Build DGEList and attach CQN offset ────────────────────────────────────
library(limma)
invisible(synLogin(silent=TRUE))
dge_cqn <- DGEList(counts = counts)
dge_cqn <- calcNormFactors(dge_cqn, method = "none")  # skip TMM; CQN handles bias
dge_cqn$offset <- cqn_offsets
dge_cqn$E <- cqn_E

dge_cqn_df <- as.data.frame(dge_cqn)
dge_cqn_row <- rownames(dge_cqn)
#dge_cqn_df[1:5,1:5]
rownames(dge_cqn_df) <- dge_cqn_row
#dge_cqn_df[1:5,1:5]

cqn_save <- list("metadata" = md_valid_qc, "counts" = counts[!zeros, ], "dge_cqn" = dge_cqn, "dge_cqn_df" = dge_cqn_df)
output_filename <- "MAYO_md_counts_cqn.rds"
saveRDS(cqn_save , file = output_filename)
synapse_file_object <- File(path = "MAYO_md_counts_cqn.rds", parent = "syn75817798")
synStore(synapse_file_object)

```


# Full data demographics (both tissues)

```{r final demo table}
# Show sample numbers for modelling by diagnosis, tissue, and sex
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)

# Prepare frequency counts wide by sex
demo_summary <- md_valid_qc %>%
  dplyr::mutate(diagnosis = factor(diagnosis, levels = c("Alzheimer Disease", "control", "pathological aging", "progressive supranuclear palsy"))) %>%
   dplyr::mutate(tissue2 = factor(tissue2, levels = c("CER", "TCX"))) %>%
  dplyr::count(tissue2, diagnosis, sex) %>%
  pivot_wider(
    names_from = sex,
    values_from = n,
    values_fill = 0
  ) %>%
  arrange(tissue2, diagnosis)

num_cols <- ncol(demo_summary %>% dplyr::select(-tissue2))

# Generate styled table
kable(
  demo_summary %>% dplyr::select(-tissue2), 
  format = "html", 
  col.names = c("Diagnosis", names(demo_summary)[-(1:2)]),
  caption = "Sample Counts by Diagnosis and Sex across Tissues"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#444444") %>%
  column_spec(1:num_cols, width = "100px") %>%
  pack_rows(index = table(demo_summary$tissue2)) # Groups rows by Tissue

```

# RNA metrics PCs for CER and TCX

Ran RAN metrics as principal components to account for technical variation


```{r RNA metrics as PCA vectors, fig.height=20, fig.width=20}
# RNA metrics as principal components / PCA of technical metrics
md_metrics <- md_valid_qc

# "X..of.chimeric.reads" dropped because all valaues were zero - all other retained
metric_cols2 <- c("picard_UNPAIRED_READS_EXAMINED"  ,        "picard_READ_PAIRS_EXAMINED"  ,    
"picard_UNMAPPED_READS"                      ,      "picard_PERCENT_DUPLICATION"                      ,
"picard_READS_UNMAPPED"                       ,     "rsem_alignable_percent"                          ,
"rsem_uniquely_aligned_percent"                ,    "samtools_reads_mapped"                           ,
"samtools_reads_duplicated"                     ,   "samtools_error_rate"                             ,
"samtools_average_length"                 ,         "samtools_average_quality"                        ,
"samtools_insert_size_average"             ,        "samtools_percentage_of_properly_paired_reads_...",
"samtools_reads_mapped_percent"             ,       "samtools_reads_mapped_and_paired_percent"        ,
"samtools_reads_properly_paired_percent"     ,      "samtools_reads_duplicated_percent"               ,
"samtools_percent_mapped_X"                   ,     "samtools_percent_mapped_Y"                       ,
"cutadapt_percent_trimmed_R2"                  ,    "cutadapt_percent_trimmed_R1"                     ,
"cutadapt_mean_percent_trimmed"                 ,   "Average.input.read.length"                       ,
"Uniquely.mapped.reads.."                        ,  "Number.of.splices..GT.AG"                        ,
"Number.of.splices..GC.AG"                        , "Number.of.splices..AT.AC"                        ,
"Mismatch.rate.per.base..."   ,                     "Deletion.rate.per.base"                          ,
"Insertion.rate.per.base"      ,                    "X..of.reads.mapped.to.multiple.loci"             ,
"X..of.reads.mapped.to.too.many.loci",              "X..of.reads.unmapped..too.many.mismatches"       ,
"X..of.reads.unmapped..too.short"     ,             "X..of.reads.unmapped..other"                     )

## adding in technical covariates ----
# RNA metrics described using PCA
## RNA metrics PCA to account for technical variability across batches
#metric_cols <- colnames(md_metrics[,c(33:69)])
#results_metrics <- compare_methods(md,metric_cols)
#sig_metrics <- results_metrics %>% filter(p_adj < 0.05) %>% arrange(p_adj) %>% pull(metric)
#cat("Significant metrics (FDR < 0.05):", length(sig_metrics), "\n")

#rownames(md) <- NULL

library(Polychrome)
p_colors <- glasbey.colors(19)[-1] #removes first color which is white
my_shapes <- c(1:19)


# ── p3: PC1 vs PC2 scores, coloured by library prep ─────────────────────────
rownames(md_metrics) <- md_metrics$specimenID
metric_cols_mat <- md_metrics[,metric_cols2]
metric_cols_mat <- as.data.frame(metric_cols_mat)
metric_cols_mat_t <- t(metric_cols_mat)

outlier_list <- list()
pca_result_outliers <- run_pca_outliers(metric_cols_mat_t,
                                         md_metrics,
                                          center = TRUE,
                                          scale = TRUE,
                                          n_pcs = 2)
pca_df <- as.data.frame(pca_result_outliers$scores[,c("specimenID", "PC1","PC2","PC3","PC4","PC5","PC6")])

p1 <- plot_pca(pca_result_outliers, "PC1", "PC2", "tissue" , title_var = c(paste("MAYO RNA metrics",sep="") ))
p2 <- plot_pca(pca_result_outliers, "PC3", "PC4", "tissue" , title_var = c(paste("MAYO RNA metrics",sep="") ))

 
p3a <- plot_pca(pca_result_outliers, "PC1", "PC2", "specimenIdSource" , title_var = c(paste("MAYO RNA metrics",sep="") ))

# ── p4: loadings — which metrics drive each PC ───────────────────────────────
p4 <- plot_pca_loadings2(pca_result_outliers)
#ggsave(paste("PCA_loadings_both_sexes_DLPFC.png",sep=""), plot =p4, width = 15, height = 10, dpi = 150)

# ── p5: scree plot ────────────────────────────────────────────────────────────
p5 <- plot_scree(pca_result_outliers)
#ggsave(paste("PCA_scree_both_sexes_DLPFC.png",sep=""), plot =p5, width = 15, height = 10, dpi = 150)

#png("~/ROSMAP_DLPFC_CN_PCC_RNA_metrics_30July2026.png", width = 1700, height = 1900)
library(patchwork)
plot_list <- list(p1,p3a,p4,p5) 
wrap_plots(plot_list, nrow = 2, ncol = 2)
#dev.off()

# metrics_pcs 
metrics_pca_df <- pca_df %>% dplyr::rename(PC1_metrics = PC1, PC2_metrics = PC2, PC3_metrics = PC3, PC4_metrics = PC4, PC5_metrics = PC5) %>% dplyr::select(specimenID, PC1_metrics, PC2_metrics, PC3_metrics, PC4_metrics, PC5_metrics)

md_valid_qc_metrics_pca <- merge(md_valid_qc, metrics_pca_df, by="specimenID")
```

# Demographics for for gene expression models

**Final analytic dataset: `r nrow(md_valid_qc_metrics_pca)` sample(s)**

**`r nrow(counts)` genes(s)**


```{r prep of final files}
md_all <- md_valid_qc_metrics_pca
counts2 <- counts[,colnames(counts) %in% md_all$specimenID]

# 1. Create a logical vector of matching columns
keep_cols <- colnames(dge_cqn) %in% md_all$specimenID

# 2. Subset the main DGEList object
dge_cqn2 <- dge_cqn[, keep_cols]

# 3. Explicitly subset the custom $E matrix
dge_cqn2$E <- dge_cqn2$E[, keep_cols]

dge_cqn_df2 <- dge_cqn_df[,colnames(dge_cqn_df) %in% md_all$specimenID]
```



```{r FINAL FILES to synapse, include=FALSE}
invisible(synLogin(silent=TRUE))
cqn_save2 <- list("metadata" = md_all, "counts" = counts2, "dge_cqn" = dge_cqn2, "dge_cqn_df" = dge_cqn_df2)
output_filename2 <- "MAYO_md_counts_cqn_FINAL.rds"
saveRDS(cqn_save2 , file = output_filename2)
synapse_file_object2 <- File(path = "MAYO_md_counts_cqn_FINAL.rds", parent = "syn75817798")
synStore(synapse_file_object2)
```




**Volcano plots of differential expression from models using counts**

```{r volcano-grid, fig.width=10, fig.height=7, echo=FALSE}
invisible(synLogin(silent=TRUE))
file1 <- synGet("syn76748651") # DE wtih residualized counts
de <- readRDS(file1$path) 
#ls(de)
females_CER6 <- de$females_CER6
males_CER6 <- de$males_CER6
females_TCX6 <- de$females_TCX6
males_TCX6 <- de$males_TCX6

plot_volcano <- function(data, 
                         title = "Volcano Plot", 
                         p_cutoff = 0.05, 
                         sig_color = "#377EB8", 
                         ns_color = "grey70",
                         xlim = c(-2, 2),
                         ylim = NULL) {
  
  p <- ggplot(data, aes(x = logFC, y = -log10(adj.P.Val))) +
    geom_point(aes(color = adj.P.Val < p_cutoff), size = 0.8, alpha = 0.6) +
    scale_color_manual(
      values = setNames(c(ns_color, sig_color), c("FALSE", "TRUE")),
      labels = c("NS", paste0("FDR < ", p_cutoff))
    ) +
    geom_hline(yintercept = -log10(p_cutoff), linetype = "dashed", color = "red") +
    labs(
      title = title, 
      x = "log2 Fold Change", 
      y = "-log10(FDR)",
      color = NULL
    ) +
    coord_cartesian(xlim = xlim, ylim = ylim) +
    theme_bw(base_size = 11)
  
  return(p)
}

females_CER6_plot <- plot_volcano(data = females_CER6, title = "MAYO, CER, female: Volcano plot")
males_CER6_plot <- plot_volcano(data = males_CER6,title = "MAYO, CER, male: Volcano plot")
females_TCX6_plot <- plot_volcano(data = females_TCX6, title = "MAYO, TCX, female: Volcano plot")
males_TCX6_plot <- plot_volcano(data = males_TCX6, title = "MAYO, TCX, male: Volcano plot")


library(patchwork)
plot_list <- list(females_CER6_plot,males_CER6_plot,
                  females_TCX6_plot,males_TCX6_plot) 
wrap_plots(plot_list, nrow = 2, ncol = 2)

```
